The Role of Variability in Learning Generalization: A Computational Modeling Approach

### 2024-05-19 Presentation/slides.qmd
some slide updates

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
subtitle: "Dissertation Defense"
format: 
  html: default
  clean-revealjs:
    #theme: moon #white
    output-ext: slide.html
    slideNumber: true
    center: true
    chalkboard: true
  pptx:
    incremental: true
author:
  - name: Thomas Gorman
    orcid: 0000-0000-0000-0000
    affiliations: Indiana University
---

## Introduction
- Investigate effects of training variability on learning and generalization
- Focus areas: visuomotor skill learning and function learning tasks

## Background and Motivation
- Importance of training variability
- Research gaps in visuomotor skill learning
- Previous research in category learning, language learning, and function learning

## Theoretical Frameworks
- Schema Theory (Schmidt, 1975)
- Desirable Difficulties Framework (Bjork & Bjork, 2011)
- Challenge Point Framework (Guadagnoli & Lee, 2004)

@@

## Overview of Current Work
- Project 1: Visuomotor projectile launching task
- Project 2: Visuomotor extrapolation task

## Project 1 - Hit The Target (HTT)
- Influence of varied practice in a projectile launching task
- Experiments 1 & 2:
  - Constant vs. varied training groups
  - Performance metrics and transfer testing

## Project 1 - Methodology
- Task description and experimental design
- Participants and procedures
- IGAS model introduction

@@

## Project 1 - Results and Discussion
- Varied training group's superior performance
- Computational modeling results with IGAS
- Theoretical implications

## Project 2 - Hit The Wall (HTW)
- Influence of varied practice in a function learning task
- Experiments 1, 2, and 3:
  - Training regimes and testing conditions
  - Learning, discrimination, and extrapolation performance metrics

## Project 2 - Methodology
- Task description and experimental design
- Participants and procedures
- Application of ALM and EXAM models

@@

## Project 2 - Results and Discussion
- Training variability's mixed results
- Modeling analysis with ALM and EXAM
- Theoretical implications

@@

## General Discussion
- Compare HTT and HTW
- Empirical findings summary
- Computational modeling contributions

## Theoretical Contributions
- Insights into training variability effects on learning and generalization
- Computational models applications

## Practical Implications
- Training programs design implications
- Future research considerations

## Limitations and Future Directions
- Study limitations
- Future research suggestions

## Conclusion
- Recap of key findings
- Final thoughts on training variability's role in learning

## Questions and Discussion
- Open floor for questions from the committee and audience

>>>>

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
subtitle: "Dissertation Defense"
format: 
  html: 
    toc: true
    #shift-heading-level-by: 1
    #toc-expand: true
  clean-revealjs:
    #theme: moon #white
    #output-ext: revealjs.html
    output-file: slides_revealjs
    slideNumber: true
    center: true
    chalkboard: true
  pptx:
    incremental: true
author:
  - name: Thomas Gorman
    url: www.tegorman13.github.io
    orcid: 0000-0001-5366-5442
    affiliations: Indiana University
---

:::: {.content-visible when-format="html" unless-format="revealjs"}
\
\
\
\
\
\
\
\
\
\

# Slides {visibility="hidden" .unnumbered .unlisted}

Presentation slides are available below. You can navigate these sides using the <kbd>←</kbd> and <kbd>→</kbd> keys. Press <kbd>m</kbd> to display all commands (e.g., press <kbd>o</kbd> to switch to slide overview)

<p class="buttons" style="text-align:center;">

<a class="btn btn-danger" target="_blank" href="/Presentation/slides_revealjs.html"><i class="fa-solid fa-arrow-up-right-from-square"></i> Open slides in new window </a>

</p>

::::: column-page-inset-right
::: {.ratio .ratio-16x9}
<iframe src="/Presentation/slides_revealjs.html"></iframe>
:::
:::::
::::

# Introduction

-   Investigate effects of training variability on learning and generalization
-   Focus areas: visuomotor skill learning and function learning tasks

---

## Background and Motivation

-   Importance of training variability
-   Research gaps in visuomotor skill learning
-   Previous research in category learning, language learning, and function learning

---

## Theoretical Frameworks

-   Schema Theory (Schmidt, 1975)
-   Desirable Difficulties Framework (Bjork & Bjork, 2011)
-   Challenge Point Framework (Guadagnoli & Lee, 2004)

@@

---

## Overview of Current Work

-   Project 1: Visuomotor projectile launching task
-   Project 2: Visuomotor extrapolation task

---

# Project 1 - Hit The Target (HTT)

-   Influence of varied practice in a projectile launching task
-   Experiments 1 & 2:
    -   Constant vs. varied training groups
    -   Performance metrics and transfer testing

---

## Project 1 - Methodology

-   Task description and experimental design
-   Participants and procedures
-   IGAS model introduction

@@

---

## Project 1 - Results and Discussion

-   Varied training group's superior performance
-   Computational modeling results with IGAS
-   Theoretical implications

---

## Computational Model

**Equation 1:** $$ Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}}) $$

**Equation 2:**
$$ d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 } $$

---

# Project 2 - Hit The Wall (HTW)

-   Influence of varied practice in a function learning task
-   Experiments 1, 2, and 3:
    -   Training regimes and testing conditions
    -   Learning, discrimination, and extrapolation performance metrics

---

## Project 2 - Methodology

-   Task description and experimental design
-   Participants and procedures
-   Application of ALM and EXAM models

@@

---

## Project 2 - Results and Discussion

-   Training variability's mixed results
-   Modeling analysis with ALM and EXAM
-   Theoretical implications

---

@@

---

# General Discussion

-   Compare HTT and HTW
-   Empirical findings summary
-   Computational modeling contributions

---

## Comparison {.smaller}

| Dimension                                       | HTT (Project 1)                                                                                                                               | HTW (Project 2)                                                                                                |
| ----------------------------------------------- | --------------------------------------------------------------------------------------------------------------------------------------------- | -------------------------------------------------------------------------------------------------------------- |
| Task Description                                | Projectile launching to hit a target                                                                                                          | Projectile launching to hit wall at a specific velocity                                                        |
| Task Complexity                                 | More complex parabolic trajectory, both x and y velocities relevant to outcome                                                                | Simpler 1D mapping of force to outcome. Only x velocity is relevant.                                           |
| Task Space                                      | More complex: xy velocity combinations closer to the solution space may still result in worse feedback due to striking the barrier.           | Simpler: smooth, linear mapping between velocity and feedback.                                                 |
| Perceptual salience of Varied Conditions | Varied conditions (# of throwing distances) are perceptually distinct, i.e. salient differences in distance between launching box and target. | Varied conditions (# of velocity bands) are less salient - only difference is the numeral displayed on screen. |
| Testing Feedback                                | Testing always included feedback                                                                                                              | Primary testing stage had no feedback.                                                                         |
| Potential for Learning during Testing           | Limited potential for learning during testing due to feedback.                                                                                | Some potential for learning during no-feedback testing by observing ball trajectory.                           |
| Training Experience                    | Varied group gets half as much experience on any one position as the constant group.                                                          | Varied group gets 1/3 as much experience on any one velocity band as the constant group.                       |
| Testing Structure                               | Random interleaving of trained/transfer testing distances.                                                                                    | Blocked structure, separately testing trained vs extrapolation testing bands.                                  |

---

## Theoretical Contributions

-   Insights into training variability effects on learning and generalization
-   Computational models applications

---

## Practical Implications

-   Training programs design implications
-   Future research considerations

---

## Limitations and Future Directions

-   Study limitations
-   Future research suggestions

---

## Conclusion

-   Recap of key findings
-   Final thoughts on training variability's role in learning

---

## Questions and Discussion

-   Open floor for questions from the committee and audience
